Variational Monte Carlo on bosonic systems

### 2018-04-03 project_1/article/article.tex
Review local energy stub

\section{Theory}
    To model the trapped bosonic gas particles we use the potential
    \begin{align}
        v(\vf{r})
        &=
        \begin{cases}
            \half m\omega^2r^2 & (\text{S}), \\
            \half m \bigl[
                \omega^2(x^2 + y^2) + \omega_z^2z^2
            \bigr] & (\text{E}),
        \end{cases}
    \end{align}
    where we can choose between a spherical (S) or an elliptical (E) harmonic
    oscillator trap.  In the latter potential the oscillatory frequency is
    different in the $z$-direction, i.e., $\omega_z \neq \omega$.  The full
    Hamiltonian of the system is given by
    \begin{align}
        H = \sum_{i = 1}^{N}h(\vf{r}_i) + \sum_{i < j}^{N}w(\vf{r}_i, \vf{r}_j),
    \end{align}
    where the single particle one-body operator, $h(\vf{r}_i)$, is given by
    \begin{align}
        h(\vf{r}_i) = -\frac{\hbar^2}{2m}\nabla_i^2
        + v(\vf{r}_i),
    \end{align}
    for equal masses, and the two-body interaction operator, $w(\vf{r}_i,
    \vf{r}_j)$, is
    \begin{align}
        w(\vf{r}_i, \vf{r}_j)
        = \begin{cases}
            \infty & |\vf{r}_i - \vf{r}_j| \leq a, \\
            0 & |\vf{r}_i - \vf{r}_j| > a,
        \end{cases}
        \label{eq:two-body_interaction}
    \end{align}
    where $a$ is the hard sphere of the particle.  The trial wavefunction,
    $\Psi_T(\vf{r})$, we will be looking at is given by
    \begin{align}
        \Psi_T(\vf{r})
        &= \Phi_T(\vf{r})
        \prod_{j < k}^N f(a, \vf{r}_j, \vf{r}_k) \\
        &= \Biggl(
            \prod_{i = 1}^N g(\alpha, \beta, \vf{r}_i)
        \Biggr)
        \prod_{j < k}^N f(a, \vf{r}_j, \vf{r}_k),
        \label{eq:initial_trial_wavefunction}
    \end{align}
    where $\alpha$ is a variational parameter and
    \begin{align}
        \vf{r} = (\vf{r}_1, \vf{r}_2, \dots, \vf{r}_N, \alpha, \beta).
    \end{align}
    Here $g$ are the single particle wavefunctions given by
    \begin{align}
        g(\alpha, \beta, \vf{r}_i)
        = \exp\bigl[
            -\alpha(x_i^2 + y_i^2 + \beta z_i^2)
        \bigr] \equiv \phi(\vf{r}_i),
    \end{align}
    and $\Phi_T(\vf{r})$ is the \textit{Slater permanent} consisting of the $N$
    first single particle wavefunctions, and the correlation wavefunction, $f$,
    given by
    \begin{align}
        f(a, \vf{r}_j, \vf{r}_k)
        &=
        \begin{cases}
            0 & |\vf{r}_j - \vf{r}_k| \leq a, \\
            \Bigl(
                1 - \frac{a}{|\vf{r}_j - \vf{r}_k|}
            \Bigr) & |\vf{r}_j - \vf{r}_k| > a.
        \end{cases}
        \label{eq:correlation_wavefunction}
    \end{align}
    We will for brevity use the notation $\phi(\vf{r}_i) = \phi_i$ and $r_{jk} =
    |\vf{r}_j - \vf{r}_k|$. In practice \autoref{eq:two-body_interaction} is
    unnecessary as the situtation where $r_{jk} \leq a$ automatically yields a
    probability density of zero from \autoref{eq:correlation_wavefunction} thus
    rejecting the state completely.

    \subsection{Local energy}
        As the many-body wavefunction creates a very large configuration space,
        where much of the wavefunction is small, we use the Metropolis algorithm
        in order to move towards regions in configuration space with
        ``sensible'' values. We define the \textit{local energy}, $E_L{\vf{r}}$,
        by
        \begin{align}
            E_L(\vf{r})
            &= \frac{H\Psi_T(\vf{r})}{\Psi_T(\vf{r})}.
        \end{align}
        If $\Psi_T(\vf{r})$ is an exact eigenfunction of the Hamiltonian, $E_L$
        will be constant. The closer $\Psi_T(\vf{r})$ is to the exact wave
        function, the less variation in $E_L$ as a function of $\vf{r}$ we get.
        One of the most computationally intensive parts of the VMC algorithm
        will be to compute $E_L$. We therefore find an analytical expression for
        $E_L$ in terms of the trial wavefunction.

@@

    \subsection{The drift force}
        A disadvantage in the use of the brute-force Metropolis-Hastings algorithm is
        that we might be spending much computational resources in an
        uninteresting part of configuration space. To make smarter moves we will
        use the Metropolis-Hastings algorithm with the additions of
        \emph{Importance Sampling}  (which will be discussed in due
        time). This alteration of the algorithm is dependent on the
        \emph{drift force} of the system.
        \begin{align}
            \vf{F}(\vf{r})
            &=
            \sum_{k = 1}^N
            \vf{F}_k(\vf{r})
            =
            \sum_{k = 1}^N
            \frac{2\nabla_k\Psi_T(\vf{r})}{\Psi_T(\vf{r})}.
            \label{eq:drift_force}
        \end{align}
        Using this expression we are able to move towards parts of
        configuration space where the gradient increases or decreases yielding a
        better choice of movements. We will mainly be interested in the drift
        force of a single particle $k$.

@@

\subsection{Interacting hard sphere bosons}
    Moving to the full system allowing $\beta$ to vary and setting $a
    \neq 0$ we can write the trial wavefunction as
    \begin{align}
        \Psi_T(\vf{r})
        &=
        \Phi_T(\vf{r})
        J(\vf{r}),
    \end{align}
    where $\ket{\Phi_T}$ is the same Slater permanent as in
    \autoref{eq:initial_trial_wavefunction} and $J(\vf{r})$ is the
    \textit{Jastrow factor} given by
    \begin{align}
        J(\vf{r})
        &=
        \exp\Biggl(
            \sum_{j < l}^N u(r_{jl})
        \Biggr),
    \end{align}
    where $r_{jk} = |\vf{r}_j - \vf{r}_k|$ and
    \begin{align}
        u(r_{jk}) = \ln\bigl[f(a, \vf{r}_j, \vf{r}_k)\bigr].
    \end{align}
    To further shorten the notation we will use $u_{jk} = u(r_{jk})$. Computing
    the gradient of the wavefunction we get
    \begin{align}
        \nabla_k\Psi_T(\vf{r})
        &=
        \Bigl[
            \nabla_k
            \Phi_T(\vf{r})
        \Bigr]
        J(\vf{r})
        + \Phi_T(\vf{r})
        \nabla_k J(\vf{r}).
    \end{align}
    The gradient of the Slater permament for particle $k$ is given by
    \begin{align}
        \nabla_k
        \Phi_T(\vf{r})
        &=
        \nabla_k\phi_k
        \prod_{i \neq k}^N\phi_i
        = \frac{\nabla_k\phi_k}{\phi_k}
        \Phi_T(\vf{r}).
    \end{align}
    The gradient of the Jastrow factor is given by
    \begin{align}
        \nabla_k J(\vf{r})
        &=
        J(\vf{r})
        \nabla_k\sum_{m < n}^N u_{mn} \\
        &= J(\vf{r})
        \Biggl(
            \sum_{m = 1}^{k - 1}\nabla_k u_{mk}
            \sum_{n = k + 1}^N\nabla_k u_{kn}
        \Biggr)
        \\
        &=
        J(\vf{r})
        \sum_{m \neq k}^N\nabla_k u_{km},
    \end{align}
    where the gradient of the interaction term splits the antisymmetric
    sum into two parts. As $r_{ij} = r_{ji}$ we can combine these sums
    into a single sum. This in total yields the gradient
    \begin{align}
        \nabla_k\Psi_T(\vf{r})
        &=
        \Biggl(
            \frac{\nabla_k\phi_k}{\phi_k}
            + \sum_{m \neq k}^N
            \nabla_k u_{km}
        \Biggr)
        \Psi_T(\vf{r}).
        \label{eq:gradient_full_wavefunction}
    \end{align}

    The Laplcian of the trial wavefunction is found by finding the divergence of
    \autoref{eq:gradient_full_wavefunction}.
    \begin{align}
        \nabla_k^2\Psi_T(\vf{r})
        &=
        \Biggl(
            \nabla_k\Biggl[
                \frac{\nabla_k\phi_k}{\phi_k}
            \Biggr]
            +
            \sum_{m \neq k}^N \nabla_k^2 u_{km}
        \Biggr)\Psi_T(\vf{r})
        \\
        &\qquad
        +
        \Biggl(
            \frac{\nabla_k\phi_k}{\phi_k}
            + \sum_{m \neq k}^N
            \nabla_k u_{km}
        \Biggr)^2
        \Psi_T(\vf{r}),
    \end{align}
    where the squared term came from taking the gradient of the trial
    wavefunction.  To further simplify we divide by the trial
    wavefunction. This yields
    \begin{align}
        \frac{\nabla_k^2\Psi_T(\vf{r})}{\Psi_T(\vf{r})}
        &=
        \frac{\nabla_k^2\phi_k}{\phi_k}
        + 2\frac{\nabla_k\phi_k}{\phi_k}
        \sum_{m \neq k}\nabla_k u_{km}
        \nonumber \\
        &\qquad
        + \sum_{m\neq k}^N\nabla_k^2 u_{km}
        + \Biggl(
            \sum_{m \neq k}^N\nabla_k u_{km}
        \Biggr)^2.
    \end{align}
    From here, the next step is to find the gradient and the Laplacian of
    the single particle functions, $\phi_k$, and the interaction
    functions $u_{km}$. For the single particle functions we use
    Cartesian coordinates when finding the derivatives while for
    the interaction functions we use spherical coordinates and do a
    variable substitution. Beginning with the gradient of the single
    particle functions we get
    \begin{align}
        \nabla_k\phi_k
        &=
        \nabla_k\exp\bigl[
            -\alpha(x_k^2 + y_k^2 + \beta z_k^2)
        \bigr] \\
        &=
        -2\alpha
        (x_k\vf{e}_i + y_k\vf{e}_j + \beta z_k\vf{e}_k)
        \phi_k.
    \end{align}
    Note that the subscripts on the unit vectors $\vf{e}_i$ are
    \textit{not} the same as the subscripts used for its components. The
    Laplacian yields
    \begin{align}
        \nabla_k^2\phi_k
        &=
        \Bigl[
            -2\alpha
            \big(
                d - 1 + \beta
            \bigr)
            \nonumber \\
            &\qquad
            + 4\alpha^2
            \bigl(
                x_k^2 + y_k^2 + \beta^2z_k^2
            \bigr)
        \Bigr]
        \phi_k,
    \end{align}
    with $d$ as the dimensionality of the problem.
    In order to derive the interaction functions we have to do a
    variable substitution using $r_{km} = |\vf{r}_k - \vf{r}_m|$. We can
    then rewrite the $\nabla_k$-operator as
    \begin{align}
        \nabla_k
        &=
        \nabla_k
        \dpd[]{r_{km}}{r_{km}}
        =
        \nabla_k r_{km} \dpd[]{}{r_{km}}
        \\
        &=
        \frac{\vf{r}_k - \vf{r}_m}{r_{km}}\dpd[]{}{r_{km}}.
    \end{align}
    Applying this version of the $\nabla_k$-operator to $u_{km}$ yields
    \begin{align}
        \nabla_k u_{km}
        &=
        \frac{\vf{r}_k - \vf{r}_m}{r_{km}}
        \dpd[]{u_{km}}{r_{km}}.
    \end{align}

@@

        We will start without importance sampling, that is by employing the
        simpler and more naïve "brute force" Monte Carlo sampling. Moreover, we
        will compute the value for the local energy analytically as well as with
        a numerical approach.  In the numerical approach we employ the well-know
        approximation for the second derivative,
        \begin{equation}
                f''(x) \approx \frac{f(x-h) - 2f(x) + f(x+h)}{h^2}.
        \end{equation}

@@

    \subsection{Energy per particle}
        For the non-interacting harmonic oscillator system there is no
        difference between a new ``particle'' and new dimensions
        \footnote{This is not entirely true in the case of our implementation of
        the Metropolis algorithms as we draw a single random particle and move
        all dimensions instead of iterating over all particles and all
        dimensions.}. This means that the inclusion of a new particle or
        dimension yields a linear scaling in the energy. Thus if we compute the
        energy per particle for systems of varying size we expect to get the
        same energy. This is most easily seen in \autoref{fig:initial_problem_b}
        and in the expression for the exact energy \autoref{eq:exact_energy}.
        This, however, is \textit{not} the case in the interacting system due to
        the Jastrow factor. Looking at \autoref{tab:10_interacting} and
        \autoref{tab:100_interacting} we see that multiplying the expected
        energy of theformer with $10$ yields a lower number than the latter. We
        can also see this behavior in \autoref{fig:interacting}, where the
        energy per particle for $N = \{10, 50, 100\}$ have been plotted in the
        same figure.

>>>>

\section{Theory}
    To model the trapped bosonic gas particles we use the potential
    \begin{align}
        v(\vf{r})
        &=
        \begin{cases}
            \half m\omega^2r^2 & (\text{S}), \\
            \half m \bigl[
                \omega^2(x^2 + y^2) + \omega_z^2z^2
            \bigr] & (\text{E}),
        \end{cases}
    \end{align}
    where we can choose between a spherical (S) or an elliptical (E) harmonic
    oscillator trap.  In the latter potential the oscillatory frequency is
    different in the $z$-direction, i.e., $\omega_z \neq \omega$.  The full
    Hamiltonian of the system is given by
    \begin{align}
        H = \sum_{i = 1}^{N}h(\vf{r}_i) + \sum_{i < j}^{N}w(\vf{r}_i, \vf{r}_j),
    \end{align}
    where the single particle one-body operator, $h(\vf{r}_i)$, is given by
    \begin{align}
        h(\vf{r}_i) = -\frac{\hbar^2}{2m}\nabla_i^2
        + v(\vf{r}_i),
    \end{align}
    for equal masses, and the two-body interaction operator, $w(\vf{r}_i,
    \vf{r}_j)$, is
    \begin{align}
        w(\vf{r}_i, \vf{r}_j)
        = \begin{cases}
            \infty & |\vf{r}_i - \vf{r}_j| \leq a, \\
            0 & |\vf{r}_i - \vf{r}_j| > a,
        \end{cases}
        \label{eq:two-body_interaction}
    \end{align}
    where $a$ is the hard sphere of the particle.  The trial wavefunction,
    $\Psi_T(\vf{r})$, we will be looking at is given by
    \begin{align}
        \Psi_T(\vf{r})
        &= \Phi_T(\vf{r})
        \prod_{j < k}^N f(a, \vf{r}_j, \vf{r}_k) \\
        &= \Biggl(
            \prod_{i = 1}^N g(\alpha, \beta, \vf{r}_i)
        \Biggr)
        \prod_{j < k}^N f(a, \vf{r}_j, \vf{r}_k),
        \label{eq:initial_trial_wavefunction}
    \end{align}
    where $\alpha$ is a variational parameter and
    \begin{align}
        \vf{r} = (\vf{r}_1, \vf{r}_2, \dots, \vf{r}_N, \alpha, \beta).
    \end{align}
    Here $g$ are the single particle wavefunctions given by
    \begin{align}
        g(\alpha, \beta, \vf{r}_i)
        = \exp\bigl[
            -\alpha(x_i^2 + y_i^2 + \beta z_i^2)
        \bigr] \equiv \phi(\vf{r}_i),
    \end{align}
    and $\Phi_T(\vf{r})$ is the \emph{Slater permanent} consisting of the $N$
    first single particle wavefunctions, and the correlation wavefunction, $f$,
    given by
    \begin{align}
        f(a, \vf{r}_j, \vf{r}_k)
        &=
        \begin{cases}
            0 & |\vf{r}_j - \vf{r}_k| \leq a, \\
            \Bigl(
                1 - \frac{a}{|\vf{r}_j - \vf{r}_k|}
            \Bigr) & |\vf{r}_j - \vf{r}_k| > a.
        \end{cases}
        \label{eq:correlation_wavefunction}
    \end{align}
    We will for brevity use the notation $\phi(\vf{r}_i) = \phi_i$ and $r_{jk} =
    |\vf{r}_j - \vf{r}_k|$. In practice \autoref{eq:two-body_interaction} is
    unnecessary as the situtation where $r_{jk} \leq a$ automatically yields a
    probability density of zero from \autoref{eq:correlation_wavefunction} thus
    rejecting the state completely.

    \subsection{Local energy}
        As the many-body wavefunction creates a very large configuration space,
        where much of the wavefunction is small, we use the Metropolis algorithm
        in order to move towards regions in configuration space with
        ``sensible'' values. We define the \emph{local energy}, $E_L(\vf{r})$,
        by
        \begin{align}
            E_L(\vf{r})
            &= \frac{H\Psi_T(\vf{r})}{\Psi_T(\vf{r})}.
        \end{align}
        If $\Psi_T(\vf{r})$ is an exact eigenfunction of the Hamiltonian, $E_L$
        will be constant. The closer $\Psi_T(\vf{r})$ is to the exact wave
        function, the less variation in $E_L$ as a function of $\vf{r}$ we get.
        This leads us to searching for values of $\vf{r}$ yielding the lowest
        variance in $E_L$.

@@

        One of the most computationally intensive parts of the Variational Monte
        Carlo algorithm will be to compute $E_L$. We will therefore find an
        analytical expression for $E_L$ in terms of the trial wavefunctions and
        the system we are exploring.

    \subsection{The drift force}
        A disadvantage in the use of the brute force Metropolis-Hastings
        algorithm is that we might be spending much computational resources in
        an uninteresting part of configuration space. To make smarter moves we
        will use the Metropolis-Hastings algorithm with the additions of
        \emph{Importance Sampling}  (which will be discussed in due time). This
        alteration of the algorithm is dependent on the \emph{drift force} of
        the system.
        \begin{align}
            \vf{F}(\vf{r})
            &=
            \sum_{k = 1}^N
            \vf{F}_k(\vf{r})
            =
            \sum_{k = 1}^N
            \frac{2\nabla_k\Psi_T(\vf{r})}{\Psi_T(\vf{r})}.
            \label{eq:drift_force}
        \end{align}
        Using this expression we are able to move towards parts of
        configuration space where the gradient increases or decreases yielding a
        better choice of movements. We will mainly be interested in the drift
        force of a single particle $k$.

@@

\subsection{Interacting hard sphere bosons}
    Moving to the full system allowing $\beta$ to vary and setting $a
    \neq 0$ we can write the trial wavefunction as
    \begin{align}
        \Psi_T(\vf{r})
        &=
        \Phi_T(\vf{r})
        J(\vf{r}),
    \end{align}
    where $\ket{\Phi_T}$ is the same Slater permanent as in
    \autoref{eq:initial_trial_wavefunction} and $J(\vf{r})$ is the
    \emph{Jastrow factor} given by
    \begin{align}
        J(\vf{r})
        &=
        \exp\Biggl(
            \sum_{j < l}^N u(r_{jl})
        \Biggr),
    \end{align}
    where $r_{jk} = |\vf{r}_j - \vf{r}_k|$ and
    \begin{align}
        u(r_{jk}) = \ln\bigl[f(a, \vf{r}_j, \vf{r}_k)\bigr].
    \end{align}
    To further shorten the notation we will use $u_{jk} = u(r_{jk})$. Computing
    the gradient of the wavefunction we get
    \begin{align}
        \nabla_k\Psi_T(\vf{r})
        &=
        \Bigl[
            \nabla_k
            \Phi_T(\vf{r})
        \Bigr]
        J(\vf{r})
        + \Phi_T(\vf{r})
        \nabla_k J(\vf{r}).
    \end{align}
    The gradient of the Slater permament for particle $k$ is given by
    \begin{align}
        \nabla_k
        \Phi_T(\vf{r})
        &=
        \nabla_k\phi_k
        \prod_{i \neq k}^N\phi_i
        = \frac{\nabla_k\phi_k}{\phi_k}
        \Phi_T(\vf{r}).
    \end{align}
    The gradient of the Jastrow factor is given by
    \begin{align}
        \nabla_k J(\vf{r})
        &=
        J(\vf{r})
        \nabla_k\sum_{m < n}^N u_{mn} \\
        &= J(\vf{r})
        \Biggl(
            \sum_{m = 1}^{k - 1}\nabla_k u_{mk}
            \sum_{n = k + 1}^N\nabla_k u_{kn}
        \Biggr)
        \\
        &=
        J(\vf{r})
        \sum_{m \neq k}^N\nabla_k u_{km},
    \end{align}
    where the gradient of the interaction term splits the antisymmetric
    sum into two parts. As $r_{ij} = r_{ji}$ we can combine these sums
    into a single sum. This in total yields the gradient
    \begin{align}
        \nabla_k\Psi_T(\vf{r})
        &=
        \Biggl(
            \frac{\nabla_k\phi_k}{\phi_k}
            + \sum_{m \neq k}^N
            \nabla_k u_{km}
        \Biggr)
        \Psi_T(\vf{r}).
        \label{eq:gradient_full_wavefunction}
    \end{align}

    The Laplcian of the trial wavefunction is found by finding the divergence of
    \autoref{eq:gradient_full_wavefunction}.
    \begin{align}
        \nabla_k^2\Psi_T(\vf{r})
        &=
        \Biggl(
            \nabla_k\Biggl[
                \frac{\nabla_k\phi_k}{\phi_k}
            \Biggr]
            +
            \sum_{m \neq k}^N \nabla_k^2 u_{km}
        \Biggr)\Psi_T(\vf{r})
        \\
        &\qquad
        +
        \Biggl(
            \frac{\nabla_k\phi_k}{\phi_k}
            + \sum_{m \neq k}^N
            \nabla_k u_{km}
        \Biggr)^2
        \Psi_T(\vf{r}),
    \end{align}
    where the squared term came from taking the gradient of the trial
    wavefunction.  To further simplify we divide by the trial
    wavefunction. This yields
    \begin{align}
        \frac{\nabla_k^2\Psi_T(\vf{r})}{\Psi_T(\vf{r})}
        &=
        \frac{\nabla_k^2\phi_k}{\phi_k}
        + 2\frac{\nabla_k\phi_k}{\phi_k}
        \sum_{m \neq k}\nabla_k u_{km}
        \nonumber \\
        &\qquad
        + \sum_{m\neq k}^N\nabla_k^2 u_{km}
        + \Biggl(
            \sum_{m \neq k}^N\nabla_k u_{km}
        \Biggr)^2.
    \end{align}
    From here, the next step is to find the gradient and the Laplacian of
    the single particle functions, $\phi_k$, and the interaction
    functions $u_{km}$. For the single particle functions we use
    Cartesian coordinates when finding the derivatives while for
    the interaction functions we use spherical coordinates and do a
    variable substitution. Beginning with the gradient of the single
    particle functions we get
    \begin{align}
        \nabla_k\phi_k
        &=
        \nabla_k\exp\bigl[
            -\alpha(x_k^2 + y_k^2 + \beta z_k^2)
        \bigr] \\
        &=
        -2\alpha
        (x_k\vf{e}_i + y_k\vf{e}_j + \beta z_k\vf{e}_k)
        \phi_k.
    \end{align}
    Note that the subscripts on the unit vectors $\vf{e}_i$ are
    \emph{not} the same as the subscripts used for its components. The
    Laplacian yields
    \begin{align}
        \nabla_k^2\phi_k
        &=
        \Bigl[
            -2\alpha
            \big(
                d - 1 + \beta
            \bigr)
            \nonumber \\
            &\qquad
            + 4\alpha^2
            \bigl(
                x_k^2 + y_k^2 + \beta^2z_k^2
            \bigr)
        \Bigr]
        \phi_k,
    \end{align}
    with $d$ as the dimensionality of the problem.
    In order to derive the interaction functions we have to do a
    variable substitution using $r_{km} = |\vf{r}_k - \vf{r}_m|$. We can
    then rewrite the $\nabla_k$-operator as
    \begin{align}
        \nabla_k
        &=
        \nabla_k
        \dpd[]{r_{km}}{r_{km}}
        =
        \nabla_k r_{km} \dpd[]{}{r_{km}}
        \\
        &=
        \frac{\vf{r}_k - \vf{r}_m}{r_{km}}\dpd[]{}{r_{km}}.
    \end{align}
    Applying this version of the $\nabla_k$-operator to $u_{km}$ yields
    \begin{align}
        \nabla_k u_{km}
        &=
        \frac{\vf{r}_k - \vf{r}_m}{r_{km}}
        \dpd[]{u_{km}}{r_{km}}.
    \end{align}

@@

        We will start without importance sampling, that is by employing the
        simpler and more naïve ``brute force'' Monte Carlo sampling. Moreover, we
        will compute the value for the local energy analytically as well as with
        a numerical approach.  In the numerical approach we employ the well-know
        approximation for the second derivative,
        \begin{equation}
                f''(x) \approx \frac{f(x-h) - 2f(x) + f(x+h)}{h^2}.
        \end{equation}

@@

    \subsection{Energy per particle}
        For the non-interacting harmonic oscillator system there is no
        difference between a new ``particle'' and new dimensions
        \footnote{This is not entirely true in the case of our implementation of
        the Metropolis algorithms as we draw a single random particle and move
        all dimensions instead of iterating over all particles and all
        dimensions.}. This means that the inclusion of a new particle or
        dimension yields a linear scaling in the energy. Thus if we compute the
        energy per particle for systems of varying size we expect to get the
        same energy. This is most easily seen in \autoref{fig:initial_problem_b}
        and in the expression for the exact energy \autoref{eq:exact_energy}.
        This, however, is \emph{not} the case in the interacting system due to
        the Jastrow factor. Looking at \autoref{tab:10_interacting} and
        \autoref{tab:100_interacting} we see that multiplying the expected
        energy of theformer with $10$ yields a lower number than the latter. We
        can also see this behavior in \autoref{fig:interacting}, where the
        energy per particle for $N = \{10, 50, 100\}$ have been plotted in the
        same figure.
